Shift-sensitive Power Transformations for Transforming Data to Normality

### 2023-02-11 manuscript/manuscript.Rmd
WIP

# Density plot.
.create_density_plot <- function(x, plot_theme){
  p <- ggplot2::ggplot(
    data = data.table::data.table("x"=x),
    mapping = ggplot2::aes(
      x = x))
  p <- p + plot_theme
  p <- p + ggplot2::geom_density()
  p <- p + ggplot2::theme(
    axis.title = ggplot2::element_blank(),
    axis.line = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    panel.grid = ggplot2::element_blank(),
    panel.border = ggplot2::element_blank(),
    plot.title=ggplot2::element_text(
      hjust=0.5,
      size=ggplot2::rel(1.0)))
  p <- p + ggplot2::xlim(c(-5, 5))

@@

    return(
      data.table::data.table(
        "distribution" = parameter_set$distribution,
        "method" = parameter_set$method,
        "version" = parameter_set$version,
        "d" = log10(parameter_set$d),
        "lambda" = transformer@lambda
      ))
  }

@@

```{=latex}
\begin{equation}
\left\{ \hat{k}_1, \hat{k}_2 \right\} = \argmin_{k_1, k_2} \sum_{i=1}^{m_d} \sum_{j=1}^{m_{out}} \left| \hat{\lambda}^r \left(\mathbf{X}_{i, j}; k_1, k_2 \right) - \hat{\lambda}_{0} \left(\mathbf{X}_i \right) \right|
\end{equation}
```
The double sum in right-hand side of the above equation can be considered as the loss that should be minimised.
Minimisation is performed using the subplex algorithm implemented in the *nloptr* package for R [@Rowan1990-sn; @Ypma2022-tc].

@@

@@

saveRDS(experiment_results, file = file.path(manuscript_dir, "robustness_comparison.RDS"))
```

@@

```{r outlier-values, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "CHECK. ")

  return(p)
}
# Only process data if the file has not been created.
if(!file.exists(file.path(manuscript_dir, "robustness_comparison_plot.RDS"))){

@@

  # Generate all experiments.
  experiments <- coro::collect(generate_experiment_data(
    x_normal = x_normal,
    x_right_skewed = x_right_skewed,
    x_left_skewed = x_left_skewed
  ))
  # Start cluster
  cl <- parallel::makeCluster(18L)

@@

} else {
  data <- readRDS(file.path(manuscript_dir, "robustness_comparison_plot.RDS"))
}

@@

data <- data[, list(
  "mean_lambda"=mean(lambda),
  "min_lambda"=min(lambda),
  "max_lambda"=max(lambda)),
  by=c("distribution", "method", "weight_method", "k")]

@@

@@

@@

x <- unlist(x, recursive = FALSE)
x <- lapply(seq_along(x), function(ii, x){
  x$distribution_id <- ii

  return(x)
})

@@

  # Determine residuals for Box-Cox transformations.
  transformer_bc <- suppressWarnings(power.transform::find_transformation_parameters(
    method = "box_cox",
    shift = TRUE,
    robust = TRUE,
    weight_method = "original_cosine",
    k1 = 0.89,
    k2 = 0.93))

@@

  residual <- stats::approx(
    x = residual_data$p,
    y = abs(residual_data$residual),
    xout = p_sample,
    rule = 2,
    ties = max)

  residual_bc <- data.table::data.table(
    "distribution_id" = x$distribution_id,
    "p" = p_sample,
    "residual" = residual,
    "method" = "box_cox")

  # Determine residuals for Yeo-Johnson transformations.
  transformer_yj <- suppressWarnings(power.transform::find_transformation_parameters(
    method = "yeo_johnson",
    shift = TRUE,
    robust = TRUE,
    weight_method = "original_cosine",
    k1 = 0.89,
    k2 = 0.93))

@@

  residual <- stats::approx(
    x = residual_data$p,
    y = abs(residual_data$residual),
    xout = p_sample,
    rule = 2,
    ties = max)

  residual_yj <- data.table::data.table(
    "distribution_id" = x$distribution_id,
    "p" = p_sample,
    "residual" = residual,
    "method" = "yeo_johnson")

@@

data <- lapply(
  x,
  compute_residuals)

@@

@@

p_bc <- ggplot2::ggplot(
  data = data[method == "Box-Cox"],
  mapping = ggplot2::aes(
    x = p,
    y = residual_error,
    colour = threshold))
p_bc <- p_bc + plot_theme
p_bc <- p_bc + ggplot2::geom_line()
p_bc <- p_bc + ggplot2::scale_colour_discrete(
  name = "Box-Cox threshold",
  type=list("50 %" = "#ABC6E2", "90 %" = "#779EC6", "95 %" = "#4E79A7", "99 %" = "#346394", "100 %" = "#1D4D7E"))
p_bc <- p_bc <- ggplot2::xlab("empirical probability")

p_yj <- ggplot2::ggplot(
  data = data[method == "Yeo-Johnson"],
  mapping = ggplot2::aes(
    x = p,
    y = residual_error,
    colour = threshold))
p_yj <- p_yj + plot_theme
p_yj <- p_yj + ggplot2::geom_line()
p_yj <- p_yj + ggplot2::scale_colour_discrete(
  name = "Yeo-Johnson threshold",
  type=list("50 %" = "#FFBD7D", "90 %" = "#FFA954", "95 %" = "#F28E2B", "99 %" = "#CD6B0B", "100 %" = "#A25000"))
p_yj <- p_yj <- ggplot2::xlab("empirical probability")

@@

>>>>

# Density plot.
.create_density_plot <- function(x, plot_theme, limits=c(-5, 5)){
  p <- ggplot2::ggplot(
    data = data.table::data.table("x"=x),
    mapping = ggplot2::aes(
      x = x))
  p <- p + plot_theme
  p <- p + ggplot2::geom_density()
  p <- p + ggplot2::theme(
    axis.title = ggplot2::element_blank(),
    axis.line = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    panel.grid = ggplot2::element_blank(),
    panel.border = ggplot2::element_blank(),
    plot.title=ggplot2::element_text(
      hjust=0.5,
      size=ggplot2::rel(1.0)))
  p <- p + ggplot2::xlim(limits)

@@

    return(
      data.table::data.table(
        "distribution" = parameter_set$distribution,
        "method" = parameter_set$method,
        "version" = parameter_set$version,
        "d" = log10(parameter_set$d),
        "lambda" = transformer@lambda,
        "x_0" = transformer@shift
      ))
  }

@@

```{=latex}
\begin{equation}
\left\{ \hat{k}_1, \hat{k}_2 \right\} = \argmin_{k_1, k_2} \sum_{i=1}^{m_d} \sum_{j=1}^{m_{out}} \left| \hat{\lambda}^r \left(\mathbf{X}_{i, j}; k_1, k_2 \right) - \hat{\lambda}_{0} \left(\mathbf{X}_i \right) \right|
\end{equation}
```
The double sum in right-hand side of the above equation can be considered as the loss that should be minimised.
Minimisation is performed using the subplex algorithm implemented in the `nloptr` package for R [@Rowan1990-sn; @Ypma2022-tc].

@@

if(!file.exists(file.path(manuscript_dir, "robustness_comparison.RDS"))){
  set.seed(95)

@@

  saveRDS(experiment_results, file = file.path(manuscript_dir, "robustness_comparison.RDS"))
}
```

@@

```{r, warning=FALSE, message=FALSE}
# Only process data if the file has not been created.
if(!file.exists(file.path(manuscript_dir, "robustness_comparison_plot.RDS"))){

@@

@@

}
```

```{r outlier-values_yeo-johnson, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Estimated $\\lambda$ values for \\textbf{shift-sensitive Yeo-Johnson} power transformations as a function of the fraction of outlier values. ",
  "Features were randomly drawn from centered, right-skewed and left-skewed normal distributions, and a fraction of values was randomly replaced by outlier values.",
  "The top row shows weighting methods that perform better than the non-robust version, ",
  "namely empirical probabilities weighted using step (emp. prob. (step)) and tapered cosine (emp. prob. (tap. cos.)) weighting functions. ",
  "The bottom row shows weighting methods that perform worse than the non-robust version. ",
  "Note that the range of $\\lambda$ values differs between top and bottom rows.")

data <- readRDS(file.path(manuscript_dir, "robustness_comparison_plot.RDS"))

@@

data <- data[method == "Yeo-Johnson", list(
  "mean_lambda"=mean(lambda),
  "min_lambda"=min(lambda),
  "max_lambda"=max(lambda)),
  by=c("distribution", "method", "weight_method", "k")]

@@

```{r outlier-values-box-cox, warning=FALSE, message=FALSE, fig.cap=cap}

data <- readRDS(file.path(manuscript_dir, "robustness_comparison_plot.RDS"))

cap <- paste0(
  "Estimated $\\lambda$ values for \\textbf{shift-sensitive Box-Cox} power transformations as a function of the fraction of outlier values. ",
  "Features were randomly drawn from centered, right-skewed and left-skewed normal distributions, and a fraction of values was randomly replaced by outlier values.",
  "The top row shows weighting methods that perform better than the non-robust version, ",
  "namely empirical probabilities weighted using step (emp. prob. (step)) and tapered cosine (emp. prob. (tap. cos.)) weighting functions. ",
  "The bottom row shows weighting methods that perform worse than the non-robust version. ",
  "Note that the range of $\\lambda$ values differs between top and bottom rows.")

data <- data[method == "Box-Cox", list(
  "mean_lambda"=mean(lambda),
  "min_lambda"=min(lambda),
  "max_lambda"=max(lambda)),
  by=c("distribution", "method", "weight_method", "k")]

top_limits <- c(-3, 4.5)
bottom_limits <- c(-4.01, 6.01)

p <- egg::ggarrange(
  plots=list(
    p_normal_t, p_right_t, p_left_t,
    p_normal_b, p_right_b, p_left_b),
  ncol=3,
  heights=c(1.0, 1.0),
  byrow=TRUE,
  draw=FALSE)
p

@@

if(!file.exists(file.path(manuscript_dir, "residual_plot.RDS"))){

  # Generate table of asymmetric generalised normal distribution parameters.
  n_distributions <- 1000

@@

  x <- unlist(x, recursive = FALSE)
  x <- mapply(
    function(ii, x){
      x$distribution_id <- ii

      return(x)},
    ii = seq_along(x),
    x = x,
    SIMPLIFY = FALSE)

@@

    # Determine residuals for Box-Cox transformations.
    transformer_bc <- suppressWarnings(power.transform::find_transformation_parameters(
      x = x$x,
      method = "box_cox",
      shift = TRUE,
      robust = TRUE,
      weight_method = "original_cosine",
      k1 = 0.76,
      k2 = 0.95))

@@

    residual <- stats::approx(
      x = residual_data$p,
      y = abs(residual_data$residual),
      xout = p_sample,
      rule = 2,
      ties = max)$y

    residual_bc <- data.table::data.table(
      "distribution_id" = x$distribution_id,
      "p" = p_sample,
      "residual" = residual,
      "method" = "box_cox",
      "has_outliers" = x$parameters$k > 0)

    # Determine residuals for Yeo-Johnson transformations.
    transformer_yj <- suppressWarnings(power.transform::find_transformation_parameters(
      x = x$x,
      method = "yeo_johnson",
      shift = TRUE,
      robust = TRUE,
      weight_method = "original_cosine",
      k1 = 0.89,
      k2 = 0.93))

@@

    residual <- stats::approx(
      x = residual_data$p,
      y = abs(residual_data$residual),
      xout = p_sample,
      rule = 2,
      ties = max)$y

    residual_yj <- data.table::data.table(
      "distribution_id" = x$distribution_id,
      "p" = p_sample,
      "residual" = residual,
      "method" = "yeo_johnson",
      "has_outliers" = x$parameters$k > 0)

@@

  # Compute all data in parallel.
  data <- parallel::parLapply(
    cl=cl,
    X=x,
    fun=compute_residuals)

  # data <- lapply(
  #   x,
  #   compute_residuals)

@@

} else {
  data <- readRDS(file.path(manuscript_dir, "residual_plot.RDS"))
}

outlier_free_data <- data[has_outliers == FALSE, list("residual_error"=stats::quantile(residual, probs=0.99)), by=c("p", "method")]

@@

p_bc <- ggplot2::ggplot(
  data = data[method == "Box-Cox"],
  mapping = ggplot2::aes(
    x = p,
    y = residual_error,
    colour = threshold))
p_bc <- p_bc + plot_theme
p_bc <- p_bc + ggplot2::geom_line()
p_bc <- p_bc + ggplot2::geom_line(
  data = outlier_free_data[method == "Box-Cox"],
  mapping = ggplot2::aes(
    x = p,
    y = residual_error),
  colour = "gray40",
  linetype = "dashed")
p_bc <- p_bc + ggplot2::scale_colour_discrete(
  name = "Box-Cox threshold",
  type=c("50 %" = "#ABC6E2", "90 %" = "#779EC6", "95 %" = "#4E79A7", "99 %" = "#346394", "100 %" = "#1D4D7E"))
p_bc <- p_bc + ggplot2::xlab("empirical probability")
p_bc <- p_bc + ggplot2::ylab("absolute residual error")
p_bc <- p_bc + ggplot2::coord_cartesian(
  xlim = c(0, 1),
  ylim = c(0, 1))
p_bc <- p_bc + ggplot2::theme(
  axis.title.x = ggplot2::element_blank(),
  axis.text.x = ggplot2::element_blank())

p_yj <- ggplot2::ggplot(
  data = data[method == "Yeo-Johnson"],
  mapping = ggplot2::aes(
    x = p,
    y = residual_error,
    colour = threshold))
p_yj <- p_yj + plot_theme
p_yj <- p_yj + ggplot2::geom_line()
p_yj <- p_yj + ggplot2::geom_line(
  data = outlier_free_data[method == "Yeo-Johnson"],
  mapping = ggplot2::aes(
    x = p,
    y = residual_error),
  colour = "gray40",
  linetype = "dashed")
p_yj <- p_yj + ggplot2::scale_colour_discrete(
  name = "Yeo-Johnson threshold",
  type=c("50 %" = "#FFBD7D", "90 %" = "#FFA954", "95 %" = "#F28E2B", "99 %" = "#CD6B0B", "100 %" = "#A25000"))
p_yj <- p_yj + ggplot2::xlab("empirical probability")
p_yj <- p_yj + ggplot2::ylab("absolute residual error")
p_yj <- p_yj + ggplot2::coord_cartesian(
  xlim = c(0, 1),
  ylim = c(0, 1))

p <- egg::ggarrange(
  plots=list(
    p_bc,
    p_yj),
  ncol=1,
  heights=c(1.0, 1.0),
  byrow=TRUE,
  draw=FALSE)
p
```

```{r goodness-of-fit, warning=FALSE, message=FALSE, fig.cap=cap}

if(!file.exists(file.path(manuscript_dir, "residual_plot.RDS"))) stop("Residual data for goodness-of-fit test does not exist.")

data <- readRDS(file.path(manuscript_dir, "residual_plot.RDS"))

central_width <- c(0.60, 0.70, 0.80, 0.90, 0.95, 1.00)

p_lower <- 0.50 - central_width / 2
p_upper <- 0.50 + central_width / 2

x <- data.table::copy(data)

# Clip empirical probabilities to centre.
x <- x[p >= p_lower & p <= p_upper]

# Compute mean absolute residual error per feature.
x <- x[, list("mare"=mean(residual)), by=c("distribution_id", "method")]
x <- x[, list("n"=.N), by=c("mare", "method")][order(mare, method)]
x[, "rejected":= 1.0 - cumsum(n) / sum(n)]
x <- rbind(data.table::data.table("mare"=c(0.0, 0.0), "method"=c("box_cox", "yeo_johnson"), "n"=0L, "rejected"=0.0))

p <- ggplot2::ggplot(
  data = x,
  mapping = ggplot2::aes(
    x = mare,
    y = rejected,
    colour = method))
p <- p + plot_theme
p <- p + ggplot2::geom_step()
p <- p + ggplot2::xlab("mean absolute residual error")
p <- p + paletteer::scale_color_paletteer_d(
  palette="ggthemes::Tableau_10",
  drop=FALSE)
```

@@

```{r eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}

if(!file.exists(file.path(manuscript_dir, "robustness_comparison_rightside_outlier.RDS"))){
  set.seed(95)

  # Add outliers, between 0.00 and 0.10.
  n_outliers <- 10
  outlier_fraction <- (seq_len(n_outliers) - 1)^2 /((n_outliers-1)^2 * 20)

            # Generate outlier values that are smaller than Q1 - 1.5 IQR or larger
            # than Q3 + 1.5 IQR.
            x_random <- stats::runif(n_draw, min=0.0, max=2.0)
            outlier <- q_upper + 1.5 * interquartile_range + x_random * interquartile_range

  saveRDS(experiment_results, file = file.path(manuscript_dir, "robustness_comparison_rightside_outlier.RDS"))
}
```
